---
title: "R Notebook"
output: html_notebook
---

### R bootcamp Bioinformatics - Spring 2025

Objectives of this notebook: 

- interpret if and for statements and custom functions
- play around with the Sanford Health Database


#### Branching control

- if statements do something if the conditional is met (is TRUE)
- if else statements do something if the conditional is met and something else if the conditional is not met
```{r}
# Do something if a conditional is met (evaluates to TRUE)
#  if(conditional){
#   statement
#  }

a <- 5

if(a>2 ){
  print("greater than 2")
} 


# Do something if a conditional is met and do something else if the conditional is not met
#  if (test_expression1) {
#  statement1
#  } else (test_expression2) {
#  statement2
#  }

b <- 1

if(b > 2 ){
  print("greater than 2")
} else {
  print("less than 2")
}


```


Write a set of expressions that test whether a person is eligible to run for US Congress
```{r}
# if (test_expression1) {
# statement1
# } else if ( test_expression2) {
# statement2
# } else if ( test_expression3) {
# statement3
# } else {
# statement4
# }
```




#### for loops

```{r}
# 
df <- c(1,2,3,4,5,6)
for (i in df){
  print(i)
  print("next")
}


```
#### Nested control

Combine for and if statements
```{r}
# Describe the following code to your neighbor

df <- c(1,2,3,4,5,6)

for (i in df) {
  if (i%%2==0) {
    print(paste(i, "meets the condition"))
  } else {
    print(paste(i, "does not meet the condition"))
    }
}

```
#### Let's plot everything in ggplot
```{r}
library(ggplot2)
library(dplyr)
```

```{r}
# Import the Sanford Data Collaborative dataset
sanfordData <- read.csv("Sanford_Data_Collaborative_Teaching_DataSet.csv")
```

#### Clean up the data
```{r}
# Convert BMI, AGE, AIC, Scheduled Clinic Visits, and missed Clinic Visits  column to numeric and remove NA cases
sanfordData$BMI <- as.numeric(as.character(sanfordData$BMI))
sanfordData$Age <- as.numeric(as.character(sanfordData$Age))
sanfordData$A1C <- as.numeric(as.character(sanfordData$A1C))
sanfordData$ScheduledClinicVisits <- as.numeric(as.character(sanfordData$ScheduledClinicVisits))
sanfordData$MissedClinicVisits <- as.numeric(as.character(sanfordData$MissedClinicVisits))

# Convert Hypertension, Vascular Disease, and Diabetes to Factors
sanfordData$Hypertension <- as.factor(sanfordData$Hypertension)
sanfordData$VascularDisease <- as.factor(sanfordData$VascularDisease)
sanfordData$Diabetes <- as.factor(sanfordData$Diabetes)
sanfordData$Status <- as.factor(sanfordData$Status)

#Remove any rows with values of NA
sanfordData <- sanfordData[!is.na(sanfordData$BMI),]
sanfordData <- sanfordData[!is.na(sanfordData$Age),]
sanfordData <- sanfordData[!is.na(sanfordData$ScheduledClinicVisits),]
sanfordData <- sanfordData[!is.na(sanfordData$MissedClinicVisits),]

# Remove Outliers for BMI
sanfordData <- sanfordData[sanfordData$BMI < 100,]
summary(sanfordData)
```

#### Subset the data
```{r}
# Take a random subsample of 100 rows
subsample <- sanfordData %>% sample_n(10000)
```


#### Simple regression
We can do a simple regression for continuous data
```{r}
# What is the relationship between BMI and age?
p <- ggplot(subsample, aes(Age, BMI, color = Sex))
p + geom_point()

# Is BMI significantly correlated with age?
lm0 <- lm(formula = BMI ~ Age, data=subsample)
summary(lm0)
```

```{r}
# and this is a much nicer way to view the relationship
q <- ggplot(subsample)
q + geom_smooth(mapping = aes(x= Age, y = BMI, color = Sex), level = 0.95)
```

#### Custom Function
```{r}
# Calculate the average smoking age by sex

calculate_smoking_by_age_sex <- function(data, age_breaks, age_labels) {
  # Ensure the number of labels matches the number of intervals
  if (length(age_labels) != (length(age_breaks) - 1)) {
    stop("Number of age labels must be one less than the number of age breaks")
  }
  
  data %>%
    mutate(AgeGroup = cut(Age, 
                          breaks = age_breaks, 
                          labels = age_labels,
                          right = TRUE)) %>%
    group_by(AgeGroup, Sex) %>%  # Now grouped by both AgeGroup and Sex
    summarize(AverageSmokingStatus = mean(SmokingStatus, na.rm = TRUE), .groups = "drop") %>%
    arrange(AgeGroup, Sex)  # Ensure correct ordering
}

# Example usage with custom age groups
age_breaks <- c(18, 30, 40, 50, 60, 70, Inf)
age_labels <- c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")

smoking_summary <- calculate_smoking_by_age_sex(sanfordData, age_breaks, age_labels)
print(smoking_summary)


```


```{r}
ss <- ggplot(sanfordData)
ss + geom_smooth(mapping = aes(x= Age, y = SmokingStatus, color = Sex), level = 0.95)
```

```{r}
vasc.dis <- ggplot(sanfordData, aes(VascularDisease, Age, colour = Sex))
vasc.dis + geom_violin()
```
