---
title: "R Notebook"
output: html_notebook
---

### R bootcamp Bioinformatics - Spring 2025

Objectives of this notebook: 

- use the Console in RStudio
- use arithmetic and relational operators
- understand the concept of assignment and assign variables to values
- use and write chunks

#### Arithmetic operators
You can use R to do simple calculations. For example, you can enter the **expression** 4 + 4 (or 4+4). An expression is a piece of code that evaluates to a value. R is an interpreted language, meaning the expressions entered are translated into machine code. The following steps occur:

1) R recognizes the expression as an arithmetic operation 
2) R is an interpreted language, so any code we write will be translated to lower-level bytecode by the R interpreter.
3) The computation happens in the processor where the CPU loads the two numbers and the CPU executes the addition instructions. The result is stored.
4) The computed value is passed back and displayed

In the Console, write expressions using the +, -, *, / , ^, %% (modulus or remainder), and %/% (integer division)

#### Relational Operators and Logical Expressions
To compare two values, you can write **logical expressions** using **relational** operators. For example, you can write the expression 4 > 5. Relational expression evaluate to TRUE or FALSE. These are fundamental for if-statements, loops, and logical filtering

== 
!= 
>
<
<=
>=

Write relational expressions in the Console

#### Assignment
To this point, the result of every expression we've written is stored, but not in a way that we can access it for further calculations. To store the result of an expression, we must assign that expression to a variable.

For example, a = 4 + 4, is read "a is assigned to 4 + 4".

Variable names have rules:

- must start with a letter (not a number or underscore)
- is case sensitive
- cannot be an R keyword
- can be in snake_case or CamelCase or some CombinationOf_both

We could run all of our code in the Console, but it is better to write it in a chunk. The code in the Console is harder to access and edit. In the chunk below, write a series of arithmetic expressions where each expression is assigned to a variable. 

```{r}
# A chunk is where we write R code
# Anything following a hashtag is ignored by the R interpreter and is only for a human reader
# To run the code in a chunk, either click on the green play arrow in the upper right corner of the chunk or type Cmd + Return or Cntl + Enter on a PC


## What happens when you run this chunk? 
a = 4 + 4
print(a)
```

```{r}
### Assignment
a = 9^2
b = 1%%3
b = a

# After these three lines of code are run, what are the values of a and b? Explain why.
```

- On a Mac, a new chunk is added by Cmd + Option + i
- On a PC, Ctrl + Alt + i

##### Practice

1) In a new chunk, assign the results of two expressions to two variables. Use a relational operator to compare the two values

2) In the chunks below, follow the prompts in the comments
```{r}
# This line generates list of random numbers, drawn from a uniform distribution
rn_list = runif(n = 10, min = 0, max = 1)
print(rn_list)
# Given the output, describe what is happening
```

```{r}
# Use a logical statement to print numbers (i) greater than 0.5

# For each item (i) in the list of random numbers (rn_list)
for (i in rn_list) {
  # if the value i is greater then 0.5, print it
  if (# replace with logical statement) {
  print(i)
  }
}

# This will give an error if you don't replace the text in the parentheses after the if
```



