---
title: "R Notebook"
output: html_notebook
---

### R bootcamp Bioinformatics - Spring 2025

Objectives of this notebook: 

- install and load external packages
- import, format, select data for analyses
- make graphs with ggplot

#### Installing and loading external packages (aka libraries)
In the lower right, you'll see a tab called "Packages". Here you can browse what you already have installed. You can select "Install" and install the package manually. Alternatively, you can edit and run the code below.
```{r}
# list all of the required packages here
list.of.packages <- c("tidyverse", "factoextra", "ggrepel", "here")

# install packages if they are not already installed
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# load all required libraries
invisible(lapply(list.of.packages, library, character.only = TRUE))
```
Once installed, you can read the documentation as you would any other library

#### Importing data
```{r}
# Load the data files for the sample metadata, 
# log transformed imputed metabolite values,
# and chemical annotation of measured metabolites

meta_all <- read_csv(
  "Complete_Data_SampleMetadata.csv",
  show_col_types = FALSE
)

log_transformed_metabolites <- read_csv(
  "Complete_Data_LogTransformedData.csv",
  show_col_types = FALSE
)

chem_annotation <- read_csv(
  "Complete_Data_ChemMetaData.csv",
  show_col_types = FALSE
)

```

#### Viewing and manipulating data types
```{r}
summary(meta_all)
```

#### Merging datasets
We can join two datasets with the left_join() function in the dplyr package
```{r}
# Join the sample metadata with the log transformed metabolite values into one dataset
log_transformed_data <- left_join(x = meta_all,
                       y = log_transformed_metabolites,
                       by = "PARENT_SAMPLE_NAME")
```

#### Adding a column
```{r}
log_transformed_data$new_column <- "expression"
```


#### Selecting analyses
```{r}
### Let's only look samples from the GRNHS (greenhouse) ENVIORNMENT
# Write a line of code to subset the log_transformed_data so that only
# samples from the GRNHS are in the new dataframe
filtered_data <- log_transformed_data[log_transformed_data$ENVIRONMENT=="GRNHS",]
```


#### Initial data visualization
Next, we will view samples with Principal Component Analysis
```{r}
# Subset the data to remove the metadata (leave only the metabolite numbers)
# This is required for PCA plotting
metabolite_data <- filtered_data[,ncol(chem_annotation):length(filtered_data)]

# Remove rows with missing values
metabolite_data <- na.omit(metabolite_data)
# Remove constant columns
metabolite_data <- metabolite_data[, apply(metabolite_data, 2, function(x) var(x) != 0)]

# Calculate a Principal Component Analysis on the cleaned metabolite data
pca_result <- prcomp(metabolite_data, 
                     center = TRUE, 
                     scale. = TRUE)

# Plot the PCA and color it based on genotype
pca_plot <- fviz_pca_ind(pca_result,
                         geom.ind = "point",
                         geom.ind.size = 8,
                         repel = TRUE, # Use text repulsion for labels
                         title = "PCA Plot", 
                         habillage = filtered_data$GENOTYPE)#  +
            #scale_colour_manual(values = custom_colors_geno)
pca_plot

```

#### Customization
```{r}
# Define a color palette and shapes for GENOTYPE
custom_colors_geno <- c("Aspeciosa" = "#fa3860",  # Dark red
                   "Asyriaca" = "#86b5da",  # Medium blue
                   "Hybrid" = "#bd96e0")  # Purple
shape_mapping <- c("Hybrid" = 16, "Asyriaca" = 17, "Aspeciosa" = 15)


```

```{r}
pca_plot <- fviz_pca_ind(pca_result,
                         geom.ind = "point",
                         geom.ind.size = 8,
                         repel = TRUE, # Use text repulsion for labels
                         title = "PCA Plot", 
                         addEllipses = TRUE,
                         habillage = filtered_data$GENOTYPE,
                         palette = custom_colors_geno) +
            scale_colour_manual(values = custom_colors_geno)
pca_plot
?fviz_pca_ind
```


For today--- 
```{r}
## Make a histogram of two chemicals 

```

```{r}
## Make a scatterplot of two of the chemicals, colored by GENOTYPE

```

```{r}
## Make a new column in the dataset which is the ratio of the two chemicals
# Log transform that ratio

```

```{r}
## Find another plot and customize it

```


#### 